An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2020-05-15 HowtoHowtoCapon/document.tex
2019-11-02
Auto-commit

% 
% In active sonar imaging the constrast and resolution can be improved by the use of adaptive beamforming. 
% 
% Adaptive beamforming 
% A critical part in the sonar image formation is the 
% The image quality is mainly limited by beampattern of the imaging arrays.  resolution and noise suppression
% 
% The quality of sonar images 
% 
% An essential part of a sonar imaging system is the beamformer. 
% 
% 
% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The beamformer's ability to suppress interference   to focus on one pixel at the time. How well a pixel value is estimated, and thus how 
% 
% A sonar image is built one pixel at the time the 
% 
% The image quality of a sonar system will depend on the imaging arrays ability to focus in 
% 
% 
% 
% Forming images done by combining data from multiple receivers in a way that to obtain focus on one pixel at the time, and assigning the output value to the intensity of that pixel. The 
% 
% Most modern active sonar systems employ phased arrays to 
% The basic way of a sonar image was formed 
% 
% A sonar system's ability to 
% 
% Adaptive beamforming has 
% 
% The image quality 
% 
% To improve the image quality of 
% 
% In high resolution sonar imaging the 
% 
% The image 
% 
% 
% In a typical sonar imaging scenario an encoded signal is transmitted to highlight a surface of interest, and the reflected wavefield sampled using a receiver array. A image is then formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time. The principle of adjusting the arrays focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
%

@@

\subsection{Beamforming}
\ifPhdDoc
\begin{figure}[htbp]\centering
\includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[htbp]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{I_beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{I_beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{I_z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{I_z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays by
%
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{I_finalZ}
\end{align}
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske2\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
3. and the beamformer output $z$ is computed from (\ref{I_finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{I_weights}). } \label{I_mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{I_mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{I_finalR});
\item computing $\eRi\1$ in (\ref{I_weights});
\item computing the beamformer output $z$ in (\ref{I_finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

% As Fig. \ref{I_mvdr_complexity} illustrates the computation of $\eR$

As observed in Fig. \ref{I_mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{I_spatialR}) and (\ref{I_finalR}) is the greatest computational burden. We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.

@@

In Fig. \ref{I_mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
\item avoid redundant multiplications;
\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
\end{enumerate}

\ifPhdDoc
\begin{figure}[tp]\centering
\includegraphics[width=\linewidth]{gfx/buske5\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps} 
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{I_mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{I_mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied, we call it a "minimum instructions" solution, and when step 2 is left out, we call it a "minimum memory" solution. Fig. \ref{I_mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{I_mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{I_mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

@@

In Fig. \ref{I_mvdr_implementation}, we illustrate how we get around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

@@

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=\linewidth]{gfx/buske8\figPostfix.eps}
% Same caption but slightly different formatting since it is now on its own page
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.
\\[.5\baselineskip]\textbf{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{I_mvdr_complexity}.
\\[.5\baselineskip]\textbf{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{I_benchmarks}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{I_mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{I_benchmarks}
\fi%
\end{figure}

@@

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske11\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske11\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
\fi\fi
\caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU. The exact same scenario and code as in Fig. \ref{I_benchmarks} was used here. With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{I_benchmarks_boston}
\end{figure}

>>>>

% 
% In active sonar imaging the constrast and resolution can be improved by the use of adaptive beamforming. 
% 
% Adaptive beamforming 
% A critical part in the sonar image formation is the 
% The image quality is mainly limited by beam pattern of the imaging arrays.  resolution and noise suppression
% 
% The quality of sonar images 
% 
% An essential part of a sonar imaging system is the beamformer. 
% 
% 
% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The beamformer's ability to suppress interference   to focus on one pixel at the time. How well a pixel value is estimated, and thus how 
% 
% A sonar image is built one pixel at the time the 
% 
% The image quality of a sonar system will depend on the imaging arrays ability to focus in 
% 
% 
% 
% Forming images done by combining data from multiple receivers in a way that to obtain focus on one pixel at the time, and assigning the output value to the intensity of that pixel. The 
% 
% Most modern active sonar systems employ phased arrays to 
% The basic way of a sonar image was formed 
% 
% A sonar system's ability to 
% 
% Adaptive beamforming has 
% 
% The image quality 
% 
% To improve the image quality of 
% 
% In high resolution sonar imaging the 
% 
% The image 
% 
% 
% In a typical sonar imaging scenario an encoded signal is transmitted to highlight a surface of interest, and the reflected wavefield sampled using a receiver array. A image is then formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time. The principle of adjusting the arrays focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
%

@@

\subsection{Beamforming}
\ifPhdDoc
\begin{figure}[htbp]\centering
\includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[htbp]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{I_beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (\Fig{I_beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{I_z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{I_z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays by
%
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{I_finalZ}
\end{align}
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske2\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
3. and the beamformer output $z$ is computed from (\ref{I_finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{I_weights}). } \label{I_mvdr_beamforming}
\end{figure}
As summarized in \Fig{I_mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{I_finalR});
\item computing $\eRi\1$ in (\ref{I_weights});
\item computing the beamformer output $z$ in (\ref{I_finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

% As \Fig{I_mvdr_complexity} illustrates the computation of $\eR$

As observed in \Fig{I_mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{I_spatialR}) and (\ref{I_finalR}) is the greatest computational burden. We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.

@@

In \Fig{I_mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
\item avoid redundant multiplications;
\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
\end{enumerate}

\ifPhdDoc
\begin{figure}[tp]\centering
\includegraphics[width=\linewidth]{gfx/buske5\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps} 
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in \Fig{I_mvdr_complexity} (higher is better). Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{I_mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied, we call it a "minimum instructions" solution, and when step 2 is left out, we call it a "minimum memory" solution. \Fig{I_mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{I_mvdr_formulas}. Here the complexity curves are relative to the reference implementation in \Fig{I_mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

@@

In \Fig{I_mvdr_implementation}, we illustrate how we get around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

@@

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=\linewidth]{gfx/buske8\figPostfix.eps}
% Same caption but slightly different formatting since it is now on its own page
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.
\\[.5\baselineskip]\textbf{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in \Fig{I_mvdr_complexity}.
\\[.5\baselineskip]\textbf{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{I_benchmarks}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in \Fig{I_mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{I_benchmarks}
\fi%
\end{figure}

@@

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske11\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske11\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
\fi\fi
\caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU. The exact same scenario and code as in \Fig{I_benchmarks} was used here. With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{I_benchmarks_boston}
\end{figure}
